Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-07-31 writeups/paper.tex
stash draft

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\bm{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\bm{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\bm{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\bm{x}$ is permitted in hydrostatic equilibrium, but we will
assume no background flow. Physically, this assumption corresponds to a
non-rotating star, or going to the corotating frame of a rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$. This both identically enforces $\rho > 0$ and
avoids numerical issues in the $\frac{\bm{\nabla}P}{\rho}$ term if $\rho$ is
small. We also define reduced pressure $\varpi = \frac{P}{\rho}$. Then, we may
rewrite the second two equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\bm{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\overline{\Upsilon} = 1, \overline{\varpi} = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u} \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}.
%,We ignore the advective components of the material derivative and consider small
% deviations $\Upsilon' = \Upsilon - \overline{\Upsilon}, \varpi' = \varpi -
% \overline{\varpi}$ about hydrostatic equilibrium:
% \begin{subequations}\label{se:lin_homo}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u}_1 &= 0,\\
%         \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
%         \pd{\bm{u}_1}{t} + \bm{\nabla}P_1
%             + \p{1 + \Upsilon_1} g\bm{z}
%             &= 0.
%     \end{align}
% \end{subequations}
The solution in this linear regime up to undetermined amplitude $A$ is
given by \citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}e^{i\p{k_{x}x + k_{z}z - \omega t}},
        \label{eq:k0z_sign}
\end{equation}
where $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
We take
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
the \emph{Brunt-V\"ais\"al\"a frequency} to be constant and take the
incompressible limit $c_s \to \infty$ where $c_s$ is the sound speed. Other
dynamical quantities are simply related to $u'_z$ (see
e.g.\ \cite{sutherland0}).

@@

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \bm{c}_{ph} &=
                \p{k_{x}\bm{x} + k_{z}\bm{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\bm{x}
                - \p{k_{x}k_{z}\bm{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{ph} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        (e.g.~\cite{drazin,sutherland1}). Note again that $c_{gz} < 0$ for
        $k_{x}, k_{z} > 0$, so upwards-propagating IGW have $k_z < 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\bm{z}$
        direction can be computed. Since the linear solution is separable as
        $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$ averaging and time
        averaging are equivalent, so we may write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \bar{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}, \omega_(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two plane wave solutions,above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z$), and we obtain
\begin{align}
    u_{z}'&(x, z, t) ={} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upwards-propagating IGW wavetrain excited deep
in the atmosphere.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\bm{u}' \cdot \bm{\nabla}$. However, since $\abs{\bm{u}'} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement of the perturbation $\bm{\xi}' =
\frac{\bm{u}'}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \abs{\bm{\xi}'} \abs{\bm{k}} \gtrsim 1.\label{eq:nl}
\end{equation}

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for $\overline{U} = \omega / k_x$ the critical layer condition
(see \autoref{ss:crit_layer} immediately below), we obtain
$\abs{\frac{u_{x}'k_{x}}{\omega}} = \abs{\xi_x k_{x}} = \abs{\xi_z k_z} \sim 1$,
equal to the nonlinearity criterion \autoref{eq:nl}. This condition predicts
where the wave's self-acceleration alone produces a critical layer.

@@

A purely horizontal shear flow $\overline{U}(z) \bm{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. For a critical value $\omega - \overline{U}_c k_{x}
= 0$, the frequency of the linear wave in the fluid's frame of reference
vanishes and critical behavior is observed. In a linear
\citep{booker_bretherton} or viscous \citep{hazel} theory, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the shear flow cannot exceed $\overline{U}_c$ the horizontal
phase velocity of the incident wave, the critical layer must instead propagate
downwards to accommodate the incident momentum. The total horizontal momentum of
the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits_0^{L_z} \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta S = 0.
\end{equation}
We define $\Delta S$ to be the change in flux across the critical layer, or
equivalently the negative of the absorbed flux. Treating $\overline{U}(z > z_c) =
\overline{U}_c, \overline{U}(z < z_c) = 0$ gives us exactly
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta S(z_c(t)).\label{eq:zc_anal}
\end{equation}
$\Delta S(z_c)$ denotes the change in $S$ across $z_c$. For constant $\Delta S$
in time and $\rho \approx \bar{\rho}$, this has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\bar{\rho}_0c_{ph, x}}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

We nondimensionalize by taking $H = N = \bar{\rho}_0 = 1$ in
\autoref{se:nl_orig}.

@@

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{z}H
        \gg 1$; we choose $k_{z}H = 1$.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from $\bm{k}$ to
        larger wavenumbers. Since well-resolved simulations using spectral
        methods have no inherent numerical viscosity, energy will accumulate at
        grid scales in the absence of artificial dissipation. We introduce
        dissipation parameter $\nu$ used for both artificial viscosity and
        artificial diffusivity. We ensure our equations of motion continue to
        conserve horizontal momentum (see \autoref{ss:strat_impl} for details).
        Under weak forcing, we may set $\nu = 0$, since energy transfer out of
        the parent $\bm{k}$ mode is negligible.
\end{itemize}

Finally, we use initial conditions $\bm{u}(0) = \Upsilon(0) = 0, \varpi(0) =
1$ corresponding to hydrostatic equilibrium and no initial fluid motion.

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, we
replace material derivatives in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z)\Upsilon,\nonumber\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2,+ \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies \citep{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. Further details of our implementation of the
fluid equations in Dedalus are described in \autoref{ss:strat_impl}.

@@

During the simulation, we expect IGW of form \autoref{eq:uz_lin} to be
excited. The incompressibility constraint allows us to compute a $\bm{u}(x, z,
t)$ profile analytically. The amplitude of the observed IGW in $\bm{u}$ relative
to analytical solution $\bm{u}_{al}$  can be estimated with estimator
$\hat{A}_i(t)$ (subscript $i$ denotes incident wave)
\begin{equation}
    \hat{A}_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \bar{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \bar{\rho}u_{al}^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
If $\bm{u} = \bm{u}_{al}$ over $z \in [z_b, z_t], \forall t$, then $\hat{A}_i(t) =
1$. The energy norm is a traditional normalization choice such that the $\bm{u}
\propto e^{z/2H}$ is evenly weighted throughout the integration region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we
expect $\hat{A}_i(t) = 1$ when integrated between the forcing and damping zones,
i.e.\  $z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are defined in
\autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t = z_b + H$
(using $z_t = z_T - \Delta z$ instead does not change the results). The
resulting measurement of $\hat{A}_i(t)$ is shown in \autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $\hat{A}_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe great agreement with \autoref{eq:uz_lin}, and in
    particular $\hat{A}_i(t)$ asymptotes to a constant value, implying continuous
    excitation of identical IGW.}\label{fig:lin_amps}
\end{figure}

>>>>

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\bm{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\bm{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\bm{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\bm{x}$ is permitted in hydrostatic equilibrium, but we will
assume no background flow, and $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star, or going to the corotating frame of a
rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ (e.g.\ see \citep{lecoanet_anel}). This both
identically enforces $\rho > 0$ and avoids numerical issues in the
$\frac{\bm{\nabla}P}{\rho}$ term if $\rho$ is small. We also define reduced
pressure $\varpi = \frac{P}{\rho}$. Then, we may rewrite the second two
equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}'}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\bm{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon = 1, \overline{\varpi} = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u}' \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime up to undetermined amplitude $A$ is given by
\citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}e^{i\p{k_{x}x + k_{z}z - \omega t}},
        \label{eq:k0z_sign}
\end{equation}
where $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
We take
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
the \emph{Brunt-V\"ais\"al\"a frequency} to be constant in the incompressible
limit $c_s \to \infty$ where $c_s$ is the sound speed. Other dynamical
quantities are simply related to $u'_z$ (see e.g.\ \cite{sutherland0}).

@@

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{ph} &=
                \p{k_{x}\bm{x} + k_{z}\bm{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\bm{x}
                - \p{k_{x}k_{z}\bm{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{ph} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        (e.g.~\cite{drazin,sutherland1}). Note again that $c_{gz} < 0$ for
        $k_{x}, k_{z} > 0$, so upwards-propagating IGW have $k_z < 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\bm{z}$
        direction can be computed. Since the linear solution consists of terms
        separable as $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$
        averaging and time averaging are equivalent. We may then write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging. Note that indeed for an
        upwards propagating IGW $k_z < 0, k_x > 0$, we have $F > 0$.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ by replacing
\autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}, \omega(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two plane wave solutions, above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z$), and we obtain
\begin{align}
    u_{z}'&(x, z, t) ={} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upwards-propagating IGW wavetrain excited deep
in the atmosphere.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\bm{u}' \cdot \bm{\nabla}$. However, since $\abs{\bm{u}'} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement of the perturbation $\bm{\xi}' =
\frac{\bm{u}'}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \abs{\bm{\xi}'} \abs{\bm{k}} \sim \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for $\overline{U} = \overline{U}_c \equiv \omega / k_x$ the
critical layer condition (see \autoref{ss:crit_layer} immediately below), we
obtain $\abs{\frac{u_{x}'k_{x}}{\omega}} = \abs{\xi_x k_{x}} = \abs{\xi_z k_z}
\sim 1$, equal to the nonlinearity criterion \autoref{eq:nl}. This condition
predicts where the wave's self-acceleration alone produces a critical layer.

@@

A purely horizontal shear flow $\overline{U}(z) \bm{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. For a critical value $\omega - \overline{U}_c k_{x}
= 0$, the horizontal wavenumber of the linear wave in the fluid's frame of
reference vanishes and critical behavior is observed. In a linear
\citep{booker_bretherton} or viscous \citep{hazel} theory, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the mean flow cannot exceed $\overline{U}_c$ the
horizontal phase velocity of the incident wave, the critical layer must instead
propagate downwards to accommodate the incident momentum. The total horizontal
momentum of the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta S = 0.
\end{equation}
$\Delta S$ is the total injected horizontal momentum to the flow. We next make a
simple two-zone approximation $\overline{U}(z > z_c) = \overline{U}_c,
\overline{U}(z < z_c) = 0$, assuming all horizontal momentum flux goes into
critical layer propagation. Then, the conservation equation becomes simple
differential equation
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta S.\label{eq:zc_anal}
\end{equation}
For constant $\Delta S$ in time and $\rho \approx \overline{\rho}$, this has
analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\overline{\rho}_0\overline{U}_c}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

We nondimensionalize by taking $H = N = \overline{\rho}_0 = 1$ in
\autoref{se:nl_orig}.

@@

        To choose $k_z$, we note astrophysical IGWs generally satisfy $\omega
        \ll N$, or equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since
        $H \lesssim R$, we aim to study $k_{z}H \gg 1$. However, to satisfy this
        well is computationally expensive: $L_z \gg H$ is required to give waves
        ample room to grow within the simulation domain, but we want $k_{z}$ to
        be sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{z}H
        \gg 1$; we choose $k_{z}H = 1$.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from $\bm{k}$ to
        larger wavenumbers. Since well-resolved simulations using spectral
        methods have no inherent numerical viscosity, energy will accumulate at
        grid scales in the absence of artificial dissipation. We introduce
        dissipation parameter $\nu$ used for both artificial viscosity and
        artificial diffusivity. We ensure our equations of motion continue to
        conserve horizontal momentum (see Appendix~\ref{se:strat_impl} for
        details). Under weak forcing, we may set $\nu = 0$, since energy
        transfer out of the parent $\bm{k}$ mode is negligible.
\end{itemize}

Finally, we use initial conditions $\bm{u}(x, z, 0) = \Upsilon(x, z, 0) = 0,
\varpi(x, z, 0) = 1$ corresponding to hydrostatic equilibrium and no initial
fluid motion.

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, we
replace material derivatives in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z)\Upsilon,\nonumber\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2,+ \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies \citep{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. Further details of our implementation of the
fluid equations in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

During the simulation, we expect IGW of form \autoref{eq:uz_lin} to be excited.
The incompressibility constraint further yields a complete analytical solution
for flow velocities $\bm{u}_{al}'(x, z, t)$. The amplitude of the observed IGW
in the simulation field $\bm{u}$ relative to analytical solution $\bm{u}_{al}'$
over some region $z \in [z_b, z_t]$ can be estimated with estimator
$\hat{A}_i(t)$ (subscript $i$ denotes incident wave)
\begin{equation}
    \hat{A}_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}'}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}(u_{al}')^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
If $\bm{u} = \bm{u}_{al}'$, then $\hat{A}_i(t) = 1$. The energy norm is a
traditional normalization choice such that the overlap between $\bm{u},
\bm{u}_{al}' \propto e^{z/2H}$ is evenly weighted throughout the integration
region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we expect
$\hat{A}_i(t) = 1$ when integrated between the forcing and damping zones,
corresponding to choices $z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are
defined in \autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For
consistency with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t =
z_b + H$ (using $z_t = z_T - \Delta z$ just below the damping layer instead does
not change the results). The resulting measurement of $\hat{A}_i(t)$ is shown in
\autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $\hat{A}_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe great agreement with \autoref{eq:uz_lin}, and in
    particular $\hat{A}_i(t)$ asymptotes to a constant value, implying continuous
    excitation of identical IGW.}\label{fig:lin_amps}
\end{figure}
